HCL-Based Color Palettes in \proglang{R}

### 2008-10-21 pkg/inst/doc/hcl-colors.Rnw
fixed compiler warnings in colorspace.c;  tidied up hcl-colors.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@44 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\title{HCL-Based Color Palettes in \proglang{R}}
\Plaintitle{HCL-Based Color Palettes in R}
\author{Achim Zeileis\\Wirtschaftsuniversit\"at Wien
   \And Kurt Hornik\\Wirtschaftsuniversit\"at Wien}
\Plainauthor{Achim Zeileis, Kurt Hornik}
\Keywords{HCL colors, qualitative palette, sequential palette, diverging palette}

\Abstract{
  The package \pkg{vcd} provides various functions providing perceptually-based
  color palettes for coding categorical data (qualitative palettes) and numerical
  variables (sequential and diverging palettes). We illustrate how these
  functions can be employed to generate various flavours of these palettes (with
  different emphases) and how they can be plugged into \proglang{R} graphics
  functions.
}

\Address{
  Achim Zeileis, Kurt Hornik\\
  Department of Statistics and Mathematics\\
  Wirtschaftsuniversit\"at Wien\\
  Augasse 2--6\\
  A-1090 Wien, Austria\\
  E-mail: \email{Achim.Zeileis@R-project.org}, \email{Kurt.Hornik@R-project.org}\\
  URL: \url{http://statmath.wu-wien.ac.at/~zeileis/},\\
  \phantom{URL:} \url{http://statmath.wu-wien.ac.at/~hornik/}
}

@@

%\VignetteIndexEntry{HCL-Based Color Palettes in R}
%\VignetteDepends{vcd,colorspace,MASS,KernSmooth,mvtnorm,kernlab}
%\VignetteKeywords{HCL colors, qualitative palette, sequential palette, diverging palette}
%\VignettePackage{colorspace}

<<preliminaries>>=
library("vcd")
library("mvtnorm")
library("kernlab")
library("KernSmooth")
@

@@

This is a companion vignette to \cite{color:Zeileis+Hornik:2006} 
and \cite{color:Zeileis+Hornik+Murrell:2007} providing
further technical details on the construction of the palettes as well as
\proglang{R} code illustrating the use of varying palettes in practice.
The palettes as well as some graphical functions demonstrated are contained
in the package \pkg{vcd}.
As a simple convenience function we will use the function
<<pal, echo=TRUE>>=
pal <- function(col, border = "light gray", ...)
{
  n <- length(col)
  plot(0, 0, type="n", xlim = c(0, 1), ylim = c(0, 1),
    axes = FALSE, xlab = "", ylab = "", ...)
  rect(0:(n-1)/n, 0, 1:n/n, 1, col = col, border = border)
}
@
which displays a set of colors using a sequence of shaded rectangles.

In the remainder of this vignette, we first outline how different types of palettes
can be constructed and generated using the tools from \pkg{vcd}. Subsequently,
we present a collection of examples illustrating the tools in practice.

@@

Qualitative palettes are sets of colors for depicting different categories,
i.e., for coding a categorical variable. To give the same
perceptual weight to each category, chroma and luminance are kept 
and only the hue is varied for obtaining different colors
(which are consequently all balanced towards the same gray).

In \pkg{vcd}, qualitative palettes are implemented in the function
\begin{Soutput}
rainbow_hcl(n, c = 50, l = 70, start = 0, end = 360*(n-1)/n, ...)
\end{Soutput}
where \code{n} controls the number of colors in the palette. The arguments
\code{c} and \code{l} give the fixed chroma and luminance levels, respectively,
and \code{start} and \code{end} specify the range of hue angles. The function
is named after the base \proglang{R} function \code{rainbow()} which has a similar
interface but chooses colors in HSV coordinates.
Figure~\ref{fig:pal-q-examples} depicts
examples for generating qualitative sets of colors $(H, 50, 70)$ produced by
<<pal-q, echo=TRUE, eval=FALSE>>=
pal(rainbow_hcl(4, start = 30, end = 300), main = "dynamic")
pal(rainbow_hcl(4, start = 60, end = 240), main = "harmonic")
pal(rainbow_hcl(4, start = 270, end = 150), main = "cold")
pal(rainbow_hcl(4, start = 90, end = -30), main = "warm")
@
From left to right and top to down, this shows a palette from the full spectrum ($H = 30, 120, 210, 300$)
creating a `dynamic' set of colors, a `harmonic'
set with $H = 60, 120, 180, 240$, warm colors (from the blue/green part of the spectrum: 
$H = 270, 230, 190, 150$) and cold colors (from the yellow/red part of the spectrum:
$H = 90, 50, 10, 330$), respectively.

@@

Two different ``flavors'' of this formula are readily implemented in \pkg{vcd},
employing different defaults. For single hue palettes with $H_1 = H_2$ the 
function
\begin{Soutput}
sequential_hcl(n, h = 260, c = c(80, 0), l = c(30, 90), power = 1.5, ...) 
\end{Soutput}
is provided where the first element of \code{c} and \code{l} give the starting chroma and luminance
coordinate (by default colorful and dark) and the second element the ending coordinate
(by default light gray). The \code{power} argument implements the parameter $p$ 
from the $i^p$ function (and can be a vector of length 2). Sequential palettes using an
interval of hues are provided by
\begin{Soutput}
heat_hcl(n, h = c(0, 90), c = c(100, 30), l = c(50, 90), power = c(1/5, 1), ...)
\end{Soutput}
named after the HSV-based \proglang{R} function \code{heat.colors()} and
by default starts from a red and going to a yellow hue. The defaults in \code{heat_hcl()}
are set differently compared to \code{sequential_hcl()} as to make the default HCL heat
colors more similar to the HSV version. The defaults of \code{sequential_hcl()}, on the
other hand, are set as to achieve a large contrast on the luminance axis. In addition
\code{terrain_hcl()} is a wrapper for \code{heat_hcl()} producing colors similar to
the HSV-based \code{terrain.colors()} from base \proglang{R}.

>>>>

\title{HCL-Based Color Palettes in \proglang{R}}
\Plaintitle{HCL-Based Color Palettes in R}
\author{Achim Zeileis\\Wirtschaftsuniversit\"at Wien
   \And Kurt Hornik\\Wirtschaftsuniversit\"at Wien
   \And Paul Murrell\\University of Auckland}
\Plainauthor{Achim Zeileis, Kurt Hornik, Paul Murrell}
\Keywords{HCL colors, qualitative palette, sequential palette, diverging palette}

\Abstract{
  The package \pkg{colorspace} provides various functions providing perceptually-based
  color palettes for coding categorical data (qualitative palettes) and numerical
  variables (sequential and diverging palettes). We illustrate how these
  functions can be employed to generate various flavours of these palettes (with
  different emphases) and how they can be plugged into \proglang{R} graphics
  functions.
}

\Address{
  Achim Zeileis, Kurt Hornik\\
  Department of Statistics and Mathematics\\
  Wirtschaftsuniversit\"at Wien\\
  Augasse 2--6\\
  A-1090 Wien, Austria\\
  E-mail: \email{Achim.Zeileis@R-project.org}, \email{Kurt.Hornik@R-project.org}\\
  URL: \url{http://statmath.wu-wien.ac.at/~zeileis/},\\
  \phantom{URL:} \url{http://statmath.wu-wien.ac.at/~hornik/}

  Paul Murrell\\
  Department of Statistics\\
  The University of Auckland\\
  Private Bag 92019\\
  Auckland\\
  New Zealand\\
  E-mail: \email{paul@stat.auckland.ac.nz}\\
  URL: \url{http://www.stat.auckland.ac.nz/~paul/}
}

@@

%\VignetteIndexEntry{HCL-Based Color Palettes in R}
%\VignetteDepends{colorspace,vcd,MASS,KernSmooth,mvtnorm,kernlab}
%\VignetteKeywords{HCL colors, qualitative palette, sequential palette, diverging palette}
%\VignettePackage{colorspace}

<<preliminaries>>=
library("colorspace")
library("vcd")
library("mvtnorm")
library("kernlab")
library("KernSmooth")
@

@@

This is a companion vignette to \cite{color:Zeileis+Hornik:2006} 
and \cite{color:Zeileis+Hornik+Murrell:2007} providing
further technical details on the construction of the palettes as well as
\proglang{R} code illustrating the use of varying palettes in practice.
The palettes as well as some graphical functions demonstrated are contained
in the package \pkg{colorspace}.
As a simple convenience function we will use the function
<<pal, echo=TRUE>>=
pal <- function(col, border = "light gray", ...)
{
  n <- length(col)
  plot(0, 0, type="n", xlim = c(0, 1), ylim = c(0, 1),
    axes = FALSE, xlab = "", ylab = "", ...)
  rect(0:(n-1)/n, 0, 1:n/n, 1, col = col, border = border)
}
@
which displays a set of colors using a sequence of shaded rectangles.

In the remainder of this vignette, we first outline how different types of palettes
can be constructed and generated using the tools from \pkg{colorspace}. Subsequently,
we present a collection of examples illustrating the tools in practice.

@@

Qualitative palettes are sets of colors for depicting different categories,
i.e., for coding a categorical variable. To give the same
perceptual weight to each category, chroma and luminance are kept constant
and only the hue is varied for obtaining different colors
(which are consequently all balanced towards the same gray).

In \pkg{colorspace}, qualitative palettes are implemented in the function
\begin{Soutput}
rainbow_hcl(n, c = 50, l = 70, start = 0, end = 360*(n-1)/n, ...)
\end{Soutput}
where \code{n} controls the number of colors in the palette. The arguments
\code{c} and \code{l} give the fixed chroma and luminance levels, respectively,
and \code{start} and \code{end} specify the range of hue angles. The function
is named after the base \proglang{R} function \code{rainbow()} which has a similar
interface but chooses colors in HSV coordinates.
Figure~\ref{fig:pal-q-examples} depicts
examples for generating qualitative sets of colors $(H, 50, 70)$ produced by
<<pal-q, echo=TRUE, eval=FALSE>>=
pal(rainbow_hcl(4, start = 30, end = 300), main = "dynamic")
pal(rainbow_hcl(4, start = 60, end = 240), main = "harmonic")
pal(rainbow_hcl(4, start = 270, end = 150), main = "cold")
pal(rainbow_hcl(4, start = 90, end = -30), main = "warm")
@
From left to right and top to down, this shows a palette from the full spectrum ($H = 30, 120, 210, 300$)
creating a `dynamic' set of colors, a `harmonic'
set with $H = 60, 120, 180, 240$, cold colors (from the blue/green part of the spectrum: 
$H = 270, 230, 190, 150$) and warm colors (from the yellow/red part of the spectrum:
$H = 90, 50, 10, 330$), respectively.

@@

Two different ``flavors'' of this formula are readily implemented in \pkg{colorspace},
employing different defaults. For single hue palettes with $H_1 = H_2$ the 
function
\begin{Soutput}
sequential_hcl(n, h = 260, c = c(80, 0), l = c(30, 90), power = 1.5, ...) 
\end{Soutput}
is provided where the first element of \code{c} and \code{l} give the starting chroma and luminance
coordinate (by default colorful and dark) and the second element the ending coordinate
(by default gray and light). The \code{power} argument implements the parameter $p$ 
from the $i^p$ function (and can be a vector of length 2). Sequential palettes using an
interval of hues are provided by
\begin{Soutput}
heat_hcl(n, h = c(0, 90), c = c(100, 30), l = c(50, 90), power = c(1/5, 1), ...)
\end{Soutput}
named after the HSV-based \proglang{R} function \code{heat.colors()} and
by default starts from a red and going to a yellow hue. The defaults in \code{heat_hcl()}
are set differently compared to \code{sequential_hcl()} as to make the default HCL heat
colors more similar to the HSV version. The defaults of \code{sequential_hcl()}, on the
other hand, are set as to achieve a large contrast on the luminance axis. In addition
\code{terrain_hcl()} is a wrapper for \code{heat_hcl()} producing colors similar to
the HSV-based \code{terrain.colors()} from base \proglang{R}.
